Baseball Data Exploration by James Kelly
========================================================


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(dplyr)
```

#Project Summary
For this project I intend to explore the relationship between player size (height and weight) and several batting statistics (batting average and home runs). As interesting relationships emerge I will explore them in greater detail!

#Data Summary
```{r echo=FALSE, Load_the_Data}
baseball_data <- read.csv('C:/Users/jkelly/Desktop/Data Viz/baseball_data_final.csv')
dim(baseball_data)
str(baseball_data)
names(baseball_data)
```


# Data Cleaning
I intend to remove players that do not have at least one home run. This way I can focus on players that are more likely to have many at-bats at the major league level, and have demonstrated at least some ability as a power hitter.

I will then create a statistic for body mass index (BMI). I suspect that BMI will have a positive relationship with home runs, and an inverse relationship with batting average.
I expect higher BMI--> more HR, lower BMI--> more avg
```{r echo=FALSE, Data_Cleaning}
baseball_data <- subset(baseball_data, HR > 0)

baseball_data$BMI <- (baseball_data$weight * 703) / (baseball_data$height^2)
baseball_data$power <- (baseball_data$HR / baseball_data$avg)
baseball_data$log10HR <- log10(baseball_data$HR)
```


#Univariate Plots
```{r echo=FALSE, Univariate_Plots}
qplot(x = avg, data = baseball_data,
      color = I('black'), fill = I('#099DD9')) +
  ggtitle('Batting Avg')

qplot(x = HR, data = baseball_data,
      color = I('black'), fill = I('#099DD9')) +
  scale_x_log10() +
  ggtitle('Log 10 HR')

qplot(x = height, data = baseball_data,
      color = I('black'), fill = I('#099DD9')) +
  ggtitle('Height')

qplot(x = weight, data = baseball_data,
      color = I('black'), fill = I('#099DD9')) +
  ggtitle('Weight')

qplot(x = BMI, data = baseball_data, binwidth =.1,
      color = I('black'), fill = I('#099DD9')) +
  ggtitle('Batting Avg')

qplot(x = power, data = baseball_data,
      color = I('black'), fill = I('#099DD9')) +
  scale_x_log10() +
  ggtitle('Power')

```
In this section I explored the distributions for several variables. The data is, for the most part, normally distributed though their is some inherent discreteness for height and weight because the data points are integers rather than floats. This resulted in considerable discreteness for BMI. Home runs are the only variable that have a long tail and require a log10 transformation.

#Bivariate Plots
```{r echo=FALSE, Bivariate_Plots}
ggplot(aes(x = avg, y = log10(HR)),
       data = baseball_data) +
  geom_boxplot() +
  facet_wrap(~handedness) +
  ggtitle('Boxplot HR v. Batting Avg')

ggplot(aes(x = BMI, y = avg),
       data = baseball_data) +
  geom_point(position = position_jitter(w = 0.3)) +
  ylim(c(0.2, 0.3)) +
  ggtitle('Batting Average v. BMI')

cor.test(baseball_data$BMI, baseball_data$avg, method = 'pearson')

ggplot(aes(x = BMI, y = log10(HR)),
       data = baseball_data) +
  geom_point(position = position_jitter(h = .05, w = 0.3)) +
  ggtitle('Log10 HR v. BMI')

cor.test(baseball_data$BMI, log10(baseball_data$HR), method = 'pearson')

ggplot(aes(x = BMI, y = log10(power)),
       data = baseball_data) +
  geom_point(position = position_jitter(h = 0.1, w = 0.3)) +
  ylim(c(1,3.2)) +
  geom_smooth(method = 'lm') +
  ggtitle('Power v. BMI')

cor.test(baseball_data$BMI, log10(baseball_data$power), method = 'pearson')       
```

First I created a boxplot to took a look at the relationship between batting average and home runs, and faceted the data by handedness. I was surprised to learn that lefties seem to have a higher batting average and more home runs, while switch hitters struggle!

Next I looked at the relationships between BMI and batting average, and BMI and home runs. Using a scatterplot I noticed that there does not appear to be any relationship between BMI and batting average. This was confirmed through correlations analysis which revealed an r quared near zero.

There does, however, appear to be a slight positive correlation between BMI and home runs. A pearson correlation test revealed an r squared of 0.2. This is very slight, but not insignificant. 

Next, I attempted to create a metric to capture hitters' power. There are already MLB statistics that measure this, but unfortunately they were not included in the data. Instead, I added a new column to the dataframe, labeled "power" which is home runs divided by batting average. This is not a very robust statistic, but it may be helpful in measuring the relationship between BMI and hitting.

Indeed, after creating another scatterplot between BMI and "power" there does to a relationship between the two variables. A pearson test the relationship is slightly stronger than that between BMI and home runs, but the "power" variable is a bit odd so I will not pursue this much further.

Based on feedback from reviewers, I will bin BMI based on how healthy their bodyweight is.

```{r}
baseball_data$range <- ifelse(baseball_data$BMI < 25, 'healthy', 'overweight')

baseball_data$range <- ordered(baseball_data$range,
                     levels = c('healthy', 'overweight'))
summary(baseball_data$range)
```

#Multivariate Plots
```{r echo=FALSE, Multivariate_Plots}

ggplot(aes(x = BMI, y = avg),
       data = baseball_data) +
  geom_point(aes(color = handedness),
             position = position_jitter(w = 0.3)) +
  ylim(c(0.2, 0.3)) +
  geom_smooth() +
  ggtitle('Batting Average v. BMI')

ggplot(aes(x = BMI, y = log10(HR)),
       data = baseball_data) +
  geom_point(aes(color = handedness),
              position = position_jitter(h = .06, w = 0.3)) +
  geom_smooth() +
  ggtitle('Log10 HR v. BMI')

ggplot(aes(x = BMI, y = log10(power)),
       data = baseball_data) +
  geom_point(aes(color = handedness),
             position = position_jitter(h = 0.1, w = 0.3)) +
  geom_smooth() +
  ggtitle('Power v. BMI') 

fit <- lm(log10HR ~ BMI, data = baseball_data)
print(fit)

baseball_data$lm <- ((.07736 * baseball_data$BMI) - .46399)
```

In this section, I created the same plots as those in the bivariate plots section, but included the handedness variable. I had hoped the superiority of lefties and inferiority of switch hitters would be more clear, but unfortunately it was not. I will instead color the plots by the range variable.
```{r}
ggplot(aes(x = BMI, y = log10(power)),
       data = baseball_data) +
  geom_point(aes(color = range),
             position = position_jitter(h = 0.1, w = 0.3)) +
  geom_smooth() +
  ggtitle('Power v. BMI')
```



#Export to CSV
write.csv(baseball_data, file = "C:/Users/jkelly/Desktop/Data Viz/baseball_data_final.csv", row.names=FALSE)
